---
title: "Class 09"
author: "Nadia Haghani"
date: "11/1/2021"
output: pdf_document
---

## Preparing the data

First, import and read data

Read the data file
```{r}
# read.csv("WisconsinCancer.csv")
```

Save it as fna.data
```{r}
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names = 1)
```

Examine the data

```{r}
wisc.df
```
Make a new dataframe without the diagnosis or the first column.

```{r}
wisc.data <- wisc.df[,-1]
```


Set up a new vector called diagnosis that contains the diagnosis column of the original dataset

```{r}
diagnosis <- as.factor(wisc.df[,1])
```


## Exploratory data analysis

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.data)
```

> Q2. How many of the observations have a malignant diagnosis?

```{r}
length(grep("M", diagnosis))
```

OR we could look at the table

```{r}
table(diagnosis)
```

Here we see that 212 observations are malignant.

> Q3. How many variables/features in the data are suffixed with _mean?


```{r}
colnames(wisc.data)
grep("_mean", colnames(wisc.data))
length(grep("_mean", colnames(wisc.data)))
```
we see there are 10 variables suffixed with _mean.


## Performing PCA

Check to see if data needs to be scaled.

Check column means and standard deviations

```{r}
colMeans(wisc.data)
apply(wisc.data,2,sd)
```


Perform PCA on wisc.data

```{r}
wisc.pr <- prcomp(wisc.data, scale = TRUE)
summary(wisc.pr)
```


> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?


```{r}
v <- round(wisc.pr$sdev^2/sum(wisc.pr$sdev^2)*100)
v
```

Therefore, 44% of the original variance is captured by PC1.

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

4 PCs required to describe 70% variance.


> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7 PCs required to described 90% variance.




## Interpreting PCA results

Make a biplot to understand results
```{r}
biplot(wisc.pr)
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

This plot is messy and difficult to understand because all of the data is clumped together with unnecessary labels.

Instead, let's make a scatter plot of each observation along PC 1 and 2 and color the points by diagnosis



```{r}
plot(wisc.pr$x[,1:2], col = diagnosis,
     xlab = "PC1", ylab = "PC2")
```


```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,3], col = diagnosis,
     xlab = "PC1", ylab = "PC3")
```


> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?



We see that PC 2 explains more variance in the original data than PC 3. Overall, plots indicate PC1 captures the malignant vs benign samples best.



Create a data.frame for ggplot

```{r}
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis
```

Load ggplot2 package

```{r}
library(ggplot2)
```

Make a scatterplot colored by diagnosis

```{r}
ggplot(df) + 
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()
```



## Variance explained

Make scree plots showing proportion of variance explained as the number of PC increases.


Calculate variance of each component
```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```


Calculate variance explained by each PC!

```{r}
pve <- pr.var / sum(pr.var)
pve
```

Let's plot this.

```{r}
plot(pve, xlab = "Principle Component",
     ylab = "Proportion of Variance Explained",
     ylim = c(0,1), type = 'o', col = "Blue")
```


Let's look at the scree plot now.

```{r}
barplot(pve, ylab = "Percent of Variance Explained",
        names.arg=paste0("PC",1:length(pve)), las=2, axes=FALSE, col = "Blue")
axis(2, at=pve, labels=round(pve,2)*100)
```


## Communicating PCA results


> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation["concave.points_mean",1]
```



> Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

```{r}
summary(wisc.pr)
```
Therefore, 5 PCs required to explain 80% original variance.


## Hierarchical Clustering


First, scale the wisc.data data and assign the result to data.scaled

```{r}
data.scaled <- scale(wisc.data)
```

Calculate Euclidean distances between all pairs of observations and assign to data.dist

```{r}
data.dist <- dist(data.scaled)
```


Create a hierarchical clustering model using complete linkage.

```{r}
wisc.hclust <- hclust(data.dist, method = "complete")
```

> Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?


```{r}
plot(wisc.hclust)
abline(h=19, col="red", lty=2)
```


4 clusters at heigh = 19.



## Selecting number of clusters

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
```

Looking on a table

```{r}
table(wisc.hclust.clusters, diagnosis)
```

> Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

```{r}
table(cutree(wisc.hclust, k=2), diagnosis)
```

> Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

```{r}
plot(hclust(data.dist, method= "average"))
plot(hclust(data.dist, method= "single"))
plot(hclust(data.dist, method= "complete"))
plot(hclust(data.dist, method= "ward.D2"))
```

Out of all the methods, the ward.D2 yields my favorite dataset because the graph seems more fitted to the data.


## Combining methods

# Clustering on PCA results

Take PCA results and cluster
```{r}
wisc.pc.hclust<-hclust(dist(wisc.pr$x[,1:3]), method="ward.D2")
```


Plot dendrogram
```{r}
plot(wisc.pc.hclust)
abline(h=70, col="red")
```


Cut tree into k=2

```{r}
grps <- cutree(wisc.pc.hclust, k=2)
table(grps)
```

Compare diagnosis & cluster groups to eachother
```{r}
table(diagnosis, grps)
```

Plot the table data

```{r}
plot(wisc.pr$x[,1:3], col=grps)
```

Color plot by diagnosis
```{r}
plot(wisc.pr$x[,1:3], col=diagnosis)
```




Re-order the groups to make them a factor (malignant is red, benign is black)


```{r}
g <- as.factor(grps)
g <- relevel(g,2)
levels(g)
```
```

```{r}
plot(wisc.pr$x[,1:2], col=g)
```


```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]), method="ward.D2")
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```



> Q15. How well does the newly created model with four clusters separate out the two diagnoses?

```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```


> Q16. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.


```{r}
table(wisc.pr.hclust.clusters, diagnosis)
table(wisc.hclust.clusters, diagnosis)
```



We see the diagnoses separated well using the k-menas and hierarchical clustering models.


## Sensitivity/Specificity

Sensitivity refers to a test’s ability to correctly detect ill patients who do have the condition. In our example here the sensitivity is the total number of samples in the cluster identified as predominantly malignant (cancerous) divided by the total number of known malignant samples. In other words: TP/(TP+FN).

Specificity relates to a test’s ability to correctly reject healthy patients without a condition. In our example specificity is the proportion of benign (not cancerous) samples in the cluster identified as predominantly benign that are known to be benign. In other words: TN/(TN+FN).


> Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

```{r}
table(diagnosis, grps)
```

```{r}
(333+179)/nrow(wisc.data)
```

```{r}
179/(179+33)
```

```{r}
333/(24+333)
```


Using PCA to predict:

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```


Plotting:

```{r}
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q18. Which of these new patients should we prioritize for follow up based on your results?


Patient 2 in malignant group
















