---
title: "Class 17"
author: "Nadia Haghani"
date: "11/28/2021"
output: pdf_document
---

We are going to examine and compare the Covid-19 vaccination rates around San Diego.

# Getting Started

First, import vaccination data

```{r}
vax <- read.csv("Covid_data.csv")
head(vax)
```

To make the data column is useful, we can use the *lubridate* package

```{r}
# install.packages("lubridate")
library(lubridate)
today()
```



> Q1. What column details the total number of people fully vaccinated? 

column 10, persons_fully_vaccinated

> Q2. What column details the Zip code tabulation area? 

Column 2, zip_code_tabulation_area

> Q3. What is the earliest date in this dataset?

2021-01-05

> Q4. What is the latest date in this dataset?

```{r}
tail(vax)
```

2021-11-23

Call the skin function to get an overview of the dataset

```{r}
# install.packages("skimr")
skimr::skim(vax)
```

> Q5. How many numeric columns are in this dataset? 

9

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 

8355

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 

8355 / 82908 * 100
10.08% 

> Q8. [Optional]: Why might this data be missing?



## Working With Dates

We already used the lubridate function. 

```{r}
# Specify that we are using the Year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

Now we can do math with the dates

```{r}
today() - vax$as_of_date[1]
```

Using the last and the first date value we can now determine how many days the dataset span?

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```


> Q9. How many days have passed since the last update of the dataset? 

```{r}
today() - vax$as_of_date[81144]
today() - vax$as_of_date[nrow(vax)]

```


> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 

```{r}
length(unique(vax$as_of_date))
```


We can download a special tool to read zipcodes

```{r}
#install.packages("zipcodeR")
# install.packages("terra")
# library(zipcodeR) # My installation of the zipcodeR was not working
```


We can do a lot of stuff with our zipcode

```{r}
# geocode_zip('92037')
```
We can find distance between centroids of any two zipcodes in miles

```{r}
# zip_distance('92037', '92109')
```

We can also pull census data
```{r}
# reverse_zipcode(c('92037', "92109"))
```

# Focus in San Diego Area

```{r}
sd <- vax$county == "San Diego"
```

Let's use *dplyr* for a more convenient subset
```{r}
library(dplyr)
sd<- filter(vax, county == "San Diego")
nrow(sd)
```

> Q11. How many distinct zip codes are listed for San Diego County?

```{r}
length(unique(sd$zip_code_tabulation_area))
```

> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                  age12_plus_population > 10000)
which.max(sd.10$age12_plus_population)
```

```{r}
pop_university <- filter(sd, zip_code_tabulation_area == "92037")
head(pop_university)
```

```{r}
sd.date <- filter(vax, county == "San Diego" & as_of_date == "2021-11-09")
```



> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2021-11-09”? 

```{r}
mean(sd.date$percent_of_population_fully_vaccinated, na.rm = TRUE) 
```



> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2021-11-09”?

```{r}
hist(sd.date$percent_of_population_fully_vaccinated)

```


# Let's focus on just UCSD area

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

```{r}
library(ggplot2)
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x= "Date", y="Percent Vaccinated", title= "Vaccination rate for La Jolla CA")
```

## Comparing 92037 to other similar sized areas?

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2021-11-16")
nrow(vax.36)
```

> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
mean(vax.36$percent_of_population_fully_vaccinated)
```

```{r}
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x= "Date", y="Percent Vaccinated", title= "Vaccination rate for La Jolla CA")  + 
  geom_hline( yintercept = 0.6629812, col = "red", 
              alpha = 0.8, linetype= "dashed")
```


> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

> Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) +aes(percent_of_population_fully_vaccinated) + geom_histogram(bins=15)
```

Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```
Zipcode 92040 is below average
Zipcode 92109 is above average

> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)
mean(vax.36.all$percent_of_population_fully_vaccinated, na.rm=TRUE)
```

```{r}
ggplot(vax.36.all) + aes(as_of_date, percent_of_population_fully_vaccinated, group=zip_code_tabulation_area) + geom_line(alpha=0.2, color = "blue")  + labs(x="Date", y= "Percent Vaccinated", title = "Vaccination rate accross California", subtitle = "Only areas with a population above 36k are shown") + geom_hline( yintercept = 0.66,linetype= "dashed")
```
> Q21. How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?

I would hope that all classes would be virtual as many individuals are exposing themselves to their families. 



